---
title: "Economic and Health Impacts of Severe Weather Events"
author: "Philip Chase"
date: "February 22, 2015"
output: html_document
---

In the report we summarize the harm caused by severe weather in events in the US as documented by the US National Weather Service.  Analysis was performed to determine the types of weather events that caused the greatest loss of life, injury and economic harm.  The analysis showed that tornadoes are the single largest cause of fatalities causing nearly three times the fatalities then any other weather event type.  Tornadoes are also the greatest cause of injuries with twelve times the injuries than any other weather event type.  The most severe economic damage is caused by Hurricane/Typhoon. 

##Data Processing

The raw source data set for our analysis can be found at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2.  A description of the data can be found at https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

To prepare the raw data for analysis we must first download and unpack the data.

```{r cache=TRUE, message=FALSE}
# We'll need the R.utils library to extract the data from a BZ2 file
library(R.utils)

# Download data file
dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dataFile <- "stormData.csv"
dataFileBz2 <- "stormData.csv.bz2"

if (!file.exists(dataFile)){
    print("Downloading data file")
    download.file(dataURL, dataFileBz2, "curl", quiet=TRUE)
    # Extract the contents of the compressed file
    bunzip2(dataFileBz2)
}
```

Inspect the uncompressed data file

```{r}
readLines("stormData.csv",n=6)
```

The data file is a comma-separated variables with quotes around text strings.  A header row names each of the columns.

Load the uncompressed data into a data frame

```{r cache=TRUE, message=FALSE}
data <- read.csv("stormData.csv")
```

Inspect the data in the data frame

```{r cache=TRUE, message=FALSE}
str(data)
head(data, n=5)
```

###Event Types

The EVTYPE column has 985 distinct levels.

```{r}
str(data$EVTYPE)
```

The number of levels can be reduced by some basic text processing.  Here we will change the case of all values to uppercase and strip the leading and trailing space from event types.

```{r}
levels(data$EVTYPE) <- gsub("(\\w+)", "\\U\\1", levels(data$EVTYPE), perl=TRUE)
levels(data$EVTYPE) <- gsub("^\\s+", "", levels(data$EVTYPE), perl=TRUE)
levels(data$EVTYPE) <- gsub("\\s+$", "", levels(data$EVTYPE), perl=TRUE)
str(data$EVTYPE)
```

This reduces the number of levels to 890.

### Population Health Data
To address the first question, _Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?_, we will need the variables EVTYPE, FATALITIES, and INJURIES.

We can create a smaller dataset to focus on these data and then inspect it for quality issues

```{r}
popData <- data[, c("EVTYPE", "FATALITIES", "INJURIES")]
str(popData)
summary(popData)
```

There are no missing values in any variables.  The data types for each column are reasonable.

###Property Damage Data

To address the second question, _Across the United States, which types of events have the greatest economic consequences?_, we will need the variables EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP to get data on property and crop damage.

We can create a smaller dataset to focus on these data as well.

```{r}
propData <- data[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
str(propData)
summary(propData)
```

The numeric property damage and crop damage figures each have an exponent expressed in a separate character column.  This makes the values not directly comparable or combinable.  Additionally some of the exponent values are not shown in the summary above.  We need to combine the numeric and exponent data into a single numeric value to allow our analysis.  There are no missing values to interfere with this transformation.

```{r}
summary(as.factor(propData$PROPDMGEXP))
summary(as.factor(propData$CROPDMGEXP))
```

I description of the data set indicates a mapping for the text values B,M and K. Other values are not described.  David Hood and Eddie Song have indicated in the discussion fora for the Coursea Reproducible Research class that the numeric values are data entry errors.  Given the small number of occurences and small values when compared to damage in the millions and billions, these values can be ignored without significatn effect on the analysis.  I will assume this mapping:

* B,b are billions or 10^9
* M,m are millions or 10^6
* K,k are thousands or 10^3
* H,h are hundreds or 10^2
* Any numeric value, is a data entry error and will be treated is missing data
* blank should be no exponent or 10^0
* ? could be interpreted is missing data or that the coder did not understand the what an exponent was or what value to enter.  I will treat "?" as a missing value.
* "-" and "+" will also be treated as missing data.

Apply these assumptions and verify they gave the desired result

```{r}
# Make a map to replace the character values with the desired numeric exponents
expChars <- c("B", "b", "M", "m", "K", "k", "H", "h", "", "?", "-", "+", seq(0,9))
expVals <- c(9, 9, 6, 6, 3, 3, 2, 2, 0, NA, NA, NA, rep(NA,10))
expMap <- data.frame(expChars, expVals)

# Apply the map to the Property Damage Exponent to replace non-numeric values
propData$PROPDMGEXPRevised <-
  unlist(
    lapply(
      as.character(propData$PROPDMGEXP),
      function(x) {
        x <- ifelse(x %in% expMap$expChars,
                    expMap$expVals[expMap$expChars == x],
                    x) ; x
        }
      )
    )

# compare the input and output of the preceeding transformation
summary(as.factor(propData$PROPDMGEXP))
summary(as.factor(propData$PROPDMGEXPRevised))

# Repeat this process for the crop data
# Apply the map to the Crop Damage Exponent to replace non-numeric values
propData$CROPDMGEXPRevised <-
  unlist(
    lapply(
      as.character(propData$CROPDMGEXP),
      function(x) {
        x <- ifelse(x %in% expMap$expChars,
                    expMap$expVals[expMap$expChars == x],
                    x) ; x
        }
      )
    )

# compare the input and output of the preceeding transformation
summary(as.factor(propData$CROPDMGEXP))
summary(as.factor(propData$CROPDMGEXPRevised))
```

Now combine the un-exponentiated damage amount and damage exponent columns to create a damage column.  Do this for both crop damage and property damage.

```{r}
propData$cropDamage <- propData$CROPDMG * 10^as.numeric(propData$CROPDMGEXPRevised)
propData$propDamage <- propData$PROPDMG * 10^as.numeric(propData$PROPDMGEXPRevised)

# inspect the new damage figures
summary(propData$cropDamage)
summary(propData$propDamage)

```

We now combine the crop and property damage estimate into a single economic harm

```{r}
propData$economicHarm <- propData$cropDamage + propData$propDamage
```

##Results

###Population Harm

To determine which events are most harmful to population health group the data by event type, total the fatalities or injuries, and filter for the top 5 events.  Do this for both fatalities and injuries.

```{r message=FALSE}
library(dplyr)

fatalities_by_eventType <- popData %>%
  group_by(EVTYPE) %>%
  summarise(total=sum(FATALITIES)) %>%
  arrange(desc(total)) %>%
  filter(total > 600)
head(fatalities_by_eventType)

injuries_by_eventType <- popData %>%
  group_by(EVTYPE) %>%
  summarise(total=sum(INJURIES)) %>%
  arrange(desc(total)) %>%
  filter(total > 2500)
head(injuries_by_eventType)
```

Plotting the injuries and fatalities by event type shows the five most harmful event types

```{r}
library(ggplot2)

ggplot(fatalities_by_eventType, aes(x=EVTYPE, y=total)) +
  geom_bar(stat="identity") +
  ggtitle("Weather events that cause the greatest number of fatalities")  +
  xlab("Type of weather event") +
  ylab("Total fatalities")

ggplot(injuries_by_eventType, aes(x=EVTYPE, y=total)) +
  geom_bar(stat="identity") +
  ggtitle("Weather events that cause the greatest number of injuries")  +
  xlab("Type of weather event") +
  ylab("Total injuries")
```

##Economic Harm

_Across the United States, which types of events have the greatest economic consequences?_
To determine which events with the greatest economic consequences we can group the data by event type, sort by economic harm, sort by total economic harm and filter for the top 5 events.

```{r message=FALSE}
economicHarm_by_eventType <- propData %>%
  group_by(EVTYPE) %>%
  summarise(total=sum(economicHarm)) %>%
  arrange(desc(total)) %>%
  filter(total > 6000000000)
head(economicHarm_by_eventType)
```

Plotting the economic harm by event type shows the five most harmful event types

```{r}
library(ggplot2)

ggplot(economicHarm_by_eventType, aes(x=EVTYPE, y=total)) +
  geom_bar(stat="identity") +
  ggtitle("Weather events that cause the greatest economic harm")  +
  xlab("Type of weather event") +
  ylab("Total monetary loss in U.S. Dollars")
```
