---
title: "Economic and Health Impacts of Severe Weather Events"
author: "Philip Chase"
date: "February 18, 2015"
output: html_document
---

_Insert Synopsis Here_: ... there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

##Data Processing

...describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

Download and unpack the data

```{r cache=TRUE, message=FALSE}
library(R.utils)

# Download data file
dataFile <- "stormData.csv"
dataFileBz2 <- "stormData.csv.bz2"

if (!file.exists(dataFile)){
    print("Downloading data file")
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", dataFileBz2, "curl", quiet=TRUE)
    # Extract the contents of the compressed file
    bunzip2(dataFileBz2)
} else {
    print("File already exists locally")
}
```

Load the uncompressed data into a data frame

```{r cache=TRUE, message=FALSE}
data <- read.csv("stormData.csv")
```

Inspect the data

```{r cache=TRUE, message=FALSE}
str(data)
summary(data)
head(data, n=10)
```

##Results

...in which your results are presented. 

The analysis document must have at least one figure containing a plot.

Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

_Other Sections are allowed_
