---
title: "Economic and Health Impacts of Severe Weather Events"
author: "Philip Chase"
date: "February 18, 2015"
output: html_document
---

_Insert Synopsis Here_: ... there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

##Data Processing

The raw source data set for our analysis can be found at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2.  A description of the data can be found at https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

To prepare the raw data for analysis we must first download and unpack the data

```{r cache=TRUE, message=FALSE}
# We'll need the R.utils library to extract the data from a BZ2 file
library(R.utils)

# Download data file
dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dataFile <- "stormData.csv"
dataFileBz2 <- "stormData.csv.bz2"

if (!file.exists(dataFile)){
    print("Downloading data file")
    download.file(dataURL, dataFileBz2, "curl", quiet=TRUE)
    # Extract the contents of the compressed file
    bunzip2(dataFileBz2)
}
```

Inspect the uncompressed data file

```{r}
readLines("stormData.csv",n=6)
```

The data file is a comma-separated variables with quotes around text strings.  A header row names each of the columns.

Load the uncompressed data into a data frame

```{r cache=TRUE, message=FALSE}
data <- read.csv("stormData.csv")
```

Inspect the data in the data frame

```{r cache=TRUE, message=FALSE}
str(data)
head(data, n=5)
```

### Population Health Data
To address the first question, _Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?_, we will need the variables EVTYPE, FATALITIES, and INJURIES.

We can create a smaller dataset to focus on these data and then inspect it for quality issues

```{r}
popData <- data[, c("EVTYPE", "FATALITIES", "INJURIES")]
str(popData)
summary(popData)
```

There are no missing values in any variables.  The data types for each column are reasonable.

###Property Damage Data

To address the second question, _Across the United States, which types of events have the greatest economic consequences?_, we will need the variables EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP to get data on property and crop damage.

We can create a smaller dataset to focus on these data as well.

```{r}
propData <- data[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
str(propData)
summary(propData)
```

The numeric property damage and crop damage figures each have an exponent expressed in a separate character column.  This makes the values not directly comparable or combinable.  Additionally some of the exponent values are not shown in the summary above.  We need to combine the numeric and exponent data into a single numeric value to allow our analysis.  There are no missing values to interfere with this transformation.

```{r}
summary(as.factor(propData$PROPDMGEXP))
summary(as.factor(propData$CROPDMGEXP))
```

I description of the data set indicates a mapping for the text values B,M and K. Other values are not described.  David Hood and Eddie Song have indicated in the discussion fora for the Coursea Reproducible Research class that the numeric values are data entry errors.  Given the small number of occurences and small values when compared to damage in the millions and billions, these values can be ignored without significatn effect on the analysis.  I will assume this mapping:

* B,b are billions or 10^9
* M,m are millions or 10^6
* K,k are thousands or 10^3
* H,h are hundreds or 10^2
* Any numeric value, is a data entry error and will be treated is missing data
* blank should be no exponent or 10^0
* ? could be interpreted is missing data or that the coder did not understand the what an exponent was or what value to enter.  I will treat "?" as a missing value.
* "-" and "+" will also be treated as missing data.

Apply these assumptions and verify they gave the desired result

```{r}
# Make a map to replace the character values with the desired numeric exponents
expChars <- c("B", "b", "M", "m", "K", "k", "H", "h", "", "?", "-", "+", seq(0,9))
expVals <- c(9, 9, 6, 6, 3, 3, 2, 2, 0, NA, NA, NA, rep(NA,10))
expMap <- data.frame(expChars, expVals)

# Apply the map to the Property Damage Exponent to replace non-numeric values
propData$PROPDMGEXPRevised <-
  unlist(
    lapply(
      as.character(propData$PROPDMGEXP),
      function(x) {
        x <- ifelse(x %in% expMap$expChars,
                    expMap$expVals[expMap$expChars == x],
                    x) ; x
        }
      )
    )

# compare the input and output of the preceeding transformation
summary(as.factor(propData$PROPDMGEXP))
summary(as.factor(propData$PROPDMGEXPRevised))

# Repeat this process for the crop data
# Apply the map to the Crop Damage Exponent to replace non-numeric values
propData$CROPDMGEXPRevised <-
  unlist(
    lapply(
      as.character(propData$CROPDMGEXP),
      function(x) {
        x <- ifelse(x %in% expMap$expChars,
                    expMap$expVals[expMap$expChars == x],
                    x) ; x
        }
      )
    )

# compare the input and output of the preceeding transformation
summary(as.factor(propData$CROPDMGEXP))
summary(as.factor(propData$CROPDMGEXPRevised))
```

Now combine the un-exponentiated damage amount and damage exponent columns to create a damage column.  Do this for both crop damage and property damage.

```{r}
propData$cropDamage <- propData$CROPDMG * 10^as.numeric(propData$CROPDMGEXPRevised)
propData$propDamage <- propData$PROPDMG * 10^as.numeric(propData$PROPDMGEXPRevised)

# inspect the new damage figures
summary(propData$cropDamage)
summary(propData$propDamage)

```

##Results

...in which your results are presented.

The analysis document must have at least one figure containing a plot.

Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

_Other Sections are allowed_
